---
title: "Base R, Part II"
author: "Andie Creel"
output: 
  pdf_document:
    number_sections: true
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list=ls())
```

# Set up: Picking Up from Base R, Part I
In this lecture, we will talk about functions, loops and if statements. We will start basic and work up to using them with a data frame. Let's make the data frame we had constructed during the last lecture (Base R, Part I).


```{r}
# Create the data frame
myPpl <- 	data.frame(
   gender = c("Male", "non-binary","Female"), 
   male = c(T, F, F),
   height = c(152, 171.5, 165), 
   weight = c(81, 93, 78),
   age = c(42,38,26)
)

# Try referencing one column
myPpl$male # version 1
myPpl[,2] #version 2

# Try referencing one row 
myPpl[1]

# Try referencing one cell
myPpl$height[1] # version 1
myPpl[1,3] # version 2


```


# Functions 

Functions: once you have initialized them, they take in an input, perform a set of operations on them, and then give you some return value. 

**Pseudo Code Example**

- consider the function: myF(x) { y <- x + 3; return(y)}
- what does myF(3) return? 6 

Note:

- Functions are helpful when you have something that you do often 
- Rule of thumb: if you're copying and pasting code 3 times or more, make function 
- (I say if you are going to copy past ever, because even if you think it'll only be twice it'll probably be more)
- Recent example for me:
  - wrote a function to take a date and return the season
  - wrote a function to get kelvin and return Fahrenheit 


```{r}

myF <- function(x){
  y <- x - x^2
  return(y)
}

myF(.5)
myF(.25)
myF(.7)

```


# Loops

- for loops: iterates through a task for a set number of times 
- Consider these loops (psuedo code): 
  - For (i in 1 through 4) { print i }
  - For (i in 1 through 4) { print i / 4}
- Can be helpful when 
  - Iterating through a column of data and do something to each row 
  - Construct a new column and want to construct each row by scratch


```{r}

# Complicated code that is simplified by the loop
print(1)
print(2)
print(3)
print(4)

# The following loop does the exact same thing
for (i in 1:4){
  print(i)
}

# more involved 
for (i in 1:4){
  print(i/4)
}

# combining loop and function  
for (i in 1:4){
  
  y = myF(i/4)
  
  print(y)
}

# Example 

# Making a new column 
# Version one: Line by line 
myPpl$age_new_a[1] <- myPpl$age[1] + 1 
myPpl$age_new_a[2] <- myPpl$age[2] + 1 
myPpl$age_new_a[3] <- myPpl$age[3] + 1 

# Version two: loop 
for (i in 1:length(myPpl$age)) {
  myPpl$age_new_b[i] <- myPpl$age[i] + 1 # everyone aged one year
}
```


# If statements

- sometimes you want to execute a task ONLY if a certain condition is met. 
- Open the myPpl df:
  - Our RA did not record men's ages right 
  - All men are actually 3 years older than what's recorded
  - What would the correct DF look like?
- If statements let you fix a mistake like this
- Also demonstrates why the Boolean (true/false or indicator) variable is so powerful

```{r}
# goes through each row and changes age if someone is male
for (i in 1:length(myPpl$male)) {
 
   if (myPpl$male[i] == TRUE) {
    myPpl$age_new_m[i] <- myPpl$age[i] - 3
  }else{
    myPpl$age_new_m[i] <- myPpl$age[i]
  }
}

```


# Other R Tutorials


[UCLA Getting Started with R](https://stats.oarc.ucla.edu/stat/data/intro_r/intro_r_interactive_flat.html)

[git lab intro](https://rgup.gitlab.io/research_cycle/01_intro.html#class_notes)

# Some specific packages
We haven't covered packages yet, but a few good resources for the future

[ggplot](http://zevross.com/blog/2014/08/04/beautiful-plotting-in-r-a-ggplot2-cheatsheet-3/)

[dplyr and tidyr](https://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf)

<!-- [R and GIS](https://cran.r-project.org/web/views/Spatial.html) -->

<!-- [R Shiny (interactive plots)](https://shiny.rstudio.com/gallery/) -->

