---
title: "Problem Set 1, Base R: Answer Key"
author: "Andie Creel"
output: pdf_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Problem 1: Basic Operations

Create two numeric variables with any values and perform basic arithmetic operations (addition, subtraction, multiplication, division) on them.
Assign the result of a division to a new variable and print it.

```{r}
# variable one 
a <- 2

# variable two 
b <- 3

a+b
a-b
a*b

c <- a/b
c

```


# Problem 2: Working with Vectors
Create a numeric vector with at least 5 elements.
Compute the sum (`sum()`) and mean (`mean()`) of the vector.
Use a logical comparison (true/false) within a for-loop to identify which elements of the vector are greater than the mean. You may need to use the `print()` command.

```{r}
# vector
myVec <- c(1,2,3,4,5)

# sum
sum(myVec)

# mean
myMean <- mean(myVec)
myMean

# logical comparison
for (i in 1:5) {
  print((myVec[i] > myMean))
}

```


# Problem 3: Basic Data Frame Manipulation
Create a data frame with at least 3 columns and 5 rows. The columns should include a mix of numeric, character, and logical data types.
Print the entire data frame.
Print only the second column of the data frame.

```{r}
# Make DF
myDF <- 	data.frame(
   gender = c("Male", "non-binary","Female", "Male", "non-binary"), 
   male = c(T, F, F, T, F),
   height = c(152, 171.5, 165, 1222, 122)
)

# whole df 
myDF

# second column
myDF$male
myDF[,2]

```

# Problem 4: Loop and Function
Write a function that takes a numeric input and returns the square of the number.
Use a for loop to apply this function to each element of a numeric vector that you create.
Store the results in a new vector and print it using the `print()` command.

```{r}
# function 
mySquare <- function(x){
  y <- x^2
  return(y)
}

# test 
mySquare(2)

#initializing vector of NAs
newVec <- rep(NA, 5)

for (i in 1:5) {
  # using vector from before
  newVec[i] <- mySquare(myVec[i])
  
  # print
  print(newVec[i])
}

```

# Problem 5: Conditional Logic
Using the data frame created in Problem 3, write an if statement inside a loop to perform a conditional operation (e.g., if a number in a numeric column is even, replace it with half its value).
Print the modified data frame.

Practicing Googling to figure out how to code if a number is even or not.

```{r}
# loop 
for (i in 1:5) {
  
  # check if even
  if((myDF$height[i] %% 2) == 0){ 
    
     # divide by 2
    myDF$height[i] <- myDF$height[i]/2
  
  }else{
    
    # do nothing
    myDF$height[i] <- myDF$height[i] 
  }
}

```

